---
title: "Designing and fitting models: A Zombie Apocolypse"
author:
  - Jesse Wheeler
  - Qianying (Ruby) Lin
  - Spencer J. Fox
format: 
  beamer:
    classoption: "t"
    # fontsize: "11pt"
    link-citations: true
    keep_tex: true
    slide_level: 2
    section-titles: false
    aspectratio: 169
    include-in-header: "../_includes/header.tex"
    # beameroptions:
    #   - setbeamertemplate: "footline[frame number]"
    #   - setbeamertemplate: "navigation symbols{}"
    header-includes: |
       \setbeamertemplate{navigation symbols}{}
       \setbeamertemplate{footline}[page number]
    hyperrefoptions:
      - linktoc=all
  html:
    toc: true
  pdf:
    toc: true
editor_options: 
  chunk_output_type: console
bibliography: ["../sbied.bib"]
---

```{r knitr_opts,include=FALSE,cache=FALSE,purl=FALSE}
source("../_includes/setup.R", local = knitr::knit_global())

```

```{r prelims,echo=F,cache=F}
library(foreach)
library(iterators)
library(doFuture)
library(tidyverse)
library(microbenchmark)
library(pomp)
plan(multisession)

```

## Recap: Euler-Binomial Scheme

We previously considered continuous time models with a simple, one-option flow:

$$S \rightarrow E \rightarrow I \rightarrow R$$

We did this by turning continuous transition rates $\mu_{SE}$ into transition probabilities over a small time step $dt$:

$$N_{SE}(t) \sim \text{Binom}\big(S(t), 1 - e^{-\mu_{SE}\times dt}\big)$$

```{c eval=FALSE}
double p = 1 - exp(-mu_SE * dt);
double trans_SE = rbinom(S, p); // Draw from Susceptible pool
S -= trans_SE;
E += trans_SE;

```

This works perfectly when there is only **one** way out of a compartment.

## The Problem: Competing Risks

What happens when an individual can exit a compartment in multiple ways?
Consider a model where an exposed individual may become either **symptomatic** ($I$), or **asymptomatic** ($A$):

$$S \rightarrow E  \,\,\begin{matrix}  \nearrow \\ \searrow  \end{matrix}\,\begin{matrix}  A \\ \\ I  \end{matrix}\,\begin{matrix}  \searrow \\ \nearrow  \end{matrix}\,\, R$$

If we use sequential `rbinom` draws, **the first draw always gets priority**, creating a mathematical bias:

```{c eval=FALSE}
double trans_I = rbinom(E, p_I);
// Bias! Asymptomatic draw is artificially restricted by the first draw
double trans_A = rbinom(E - trans_I, p_A); 
```

## Euler-Multinomial Distribution

To avoid this bias, we need to draw both transitions at the same time. We can handle convert rates into probabilities, and do simultaneous draws, with a single distribution: **Euler-Multinomial**.

\vspace{2cm}

This is a natural extension of the process of converting continuous-time rates to probabilities, and doing binomial draws, which has two options: staying in $E$, or going to $I$. 

Now, we may have 3 options: Staying in $E$, going to $I$, or going to $A$:

$$
\big(N_{EI}(t), N_{EA}(t)) \sim \text{EulerMultinom}\big[E(t), (\mu_{EI}(t), \mu_{EA}(t))\big]
$$

## `reulermultinom` in practice

The `reulermultinom` function does this for us. It calculates the total exits based on the sum of all rates, then correctly distributes them.

```{c eval=FALSE}
// 1. Arrays for rates and output
double rates[2] = {mu_EI, mu_EA};
double trans[2];

// 2. Draw simultaneously
// (num_paths, size, &rates, dt, &trans)
reulermultinom(2, E, &rates[0], dt, &trans[0]);

// 3. Extract results
double trans_I = trans[0];
double trans_A = trans[1];
```

*Note: The `&` symbol just tells C exactly where your lists are stored in memory.*

## Introduction {.allowframebreaks}

* Our goal: practice what we learned, design a model and make predictions.
* **Scenario**: A zombie-apocalyptic infection has spread on a small island nation.
* Simulated data available on course website: weekly data for 1.5 years

```{r, echo=FALSE, fig.height=2}
zombie_df <- read.csv('raw-data/zombieCases.csv')
ggplot(zombie_df, aes(x = time, y = cases)) + 
  geom_line() + 
  geom_point() + 
  theme_bw()

```

## Goals

* Fit a model
* Estimate number of surviving humans, forecasts, etc.

# Part 1: The Scenario & Conceptual Model

## The Anomaly

* **The Setup:** Public health officials are tracking a bizarre neurological outbreak (colloquially "Zombies").
* **The Hypothesis:** The pathogen circulated completely undetected for 2 months ($t < 60$) causing no reportable symptoms. At Day 60, it mutated.
* **The Mutation:** A high (unknown) percentage of exposed individuals turn into aggressive Zombies; the remaining proportion develop a rash and become permanently immune.
* **Spread:** The virus spreads from contacts between current zombies + exposed individuals and susceptible survivors.
* **Containment:** Zombies do not die naturally. They are destroyed via mass-action combat driven by surviving humans (Susceptible + Immune).

## The Fog of War (Measurement Model)

* We only partially observe total zombies, though reporting is high.
* The exposed class (those that may become zombies) is completely invisible.
* Reporting infrastructure decays (exponentially) as the total zombie horde ($Z$) grows.

## Model building

* What compartments should we include based on what we know? 
* New infections follow similar dynamics relative to what we have seen. What about the mass-action combat destruction of zombies?
* Modeling and forecasting long periods. Should we include births / deaths?

## Independent Exercise: Map the Competing Risks (15 mins)

**On Paper or Whiteboard:**
Map the transition flow for this system. Pay special attention to competing risks (arrows leaving the same box).

1. What are the two exit pathways for a Susceptible ($S$) individual?
2. What are the three exit pathways for an Exposed ($E$) individual?
3. If $R$ holds dead bodies and destroyed zombies, do they contribute to the birth rate ($\mu_b$)?
4. Should $R$ be in the transmission denominator ($N$)? Why or why not?

# Part 2: Building the POMP Object

## Independent Exercise: Build the Model (45 mins)

**Your Task:** Work in your groups to translate your conceptual model into `pomp` C-snippets.

1. Write `rinit`, `rprocess`, `rmeasure`, and `dmeasure` snippets.
2. Build the `pomp` object.
3. Choose a reasonable set of starting parameters and use `simulate()` to ensure your model produces an outbreak curve that roughly matches the shape of the data.

# Part 3: Inference and Forecasting

## Estimation with Iterated Filtering

The dataset contains a massive gap: the first two months of the outbreak are entirely unobserved.

**The Challenge:**
When you run `mif2`, the particles must correctly estimate the early, silent transmission rate ($\beta_E$) based *only* on the magnitude of the zombie explosion that happens weeks later.

## Independent Exercise: Estimation & The Doomsday Forecast (60 mins)

1. **Estimation:** Run `mif2` to estimate the transmission rates ($\beta_E, \beta_Z$), the combat effectiveness ($\alpha$), and the immunity fraction ($\omega$). Plot your convergence traces.
2. **Forecasting:** Extract your best parameter estimates. Use `simulate()` to generate 50 forward trajectories extending 10 years into the future.
3. **Discussion:** Look at your simulated trajectories for the Susceptible ($S$) compartment. Does the human-driven combat mechanic allow humanity to bounce back, or does the dropping population trigger a runaway collapse?